Correcting for bias in high throughput sequencing data

### 2011-06-14 seqbias-paper.tex
a new models graph figure
poisson boxplot
more plotting
text revisions

\section{Motivation:}

Quantification of sequence abundance in RNA-Seq and ChiP-Seq experiments is
often conflated with protocal-specific bias, influenced by PCR amplification,
and differing primer affinities and mixtures, for example. The result is
potentially inaccurate quantification, affecting de novo gene annotation and
isoform quantification in RNA-Seq, and peak calling in ChiP-Seq.

\section{Results:}

We present a method to measure and correct for these influences using a simple
graphical model. Our model does not rely on existing gene annotations, making it
applicable to any high throughput sequencing data. We evaluate our method on
several datasets, and by multiple criteria, demonstrating that it effectively
decreases bias and increases uniformity.

@@

A significant example is the effect of transcript length on quantification.
As noted by Oshlack, et. al., transcript length biases discovery of differentially
expression, since more reads are aligned to longer transcripts, granting
higher statistical significance in tests differential expression
\cite{Oshlack2009}. Recently, some methods have been developed attempting to account for
this effect in gene set enrichment tests \cite{Young2010} \cite{Gao2011}.
Additionally, differing transcript lengths introduce ``edge
effects'', as very short fragments from the end of a transcript will not be
sequenced. NEUMA \cite{Lee2010} and FPKM \cite{Trapnell2010} are two methods
that attempt to take this into account.

The underlying nucleotide sequence also influences quantification in a manner
that is complex and protocol specific. In an ideal experiment, the number of
reads mapping to a particular position in the genome is a function of RNA
abundance and should not be additionally dependent on the sequence. As pointed
out recently, this is not the case \cite{Dohm2008} \cite{Li2010}
\cite{Hansen2010}. As illustration, Figure \ref{fig:freqs} plots this
nonuniformity in nucleotide frequencies on five data sets (see, Table
\ref{tab:datasets}), each using a different protocol.

These biases may adversely effect transcript discovery, as low level noise may
be over reported in some regions, and in others, active transcription may be
under reported. They render untrustworthy comparisons of relative abundance
between genes, and any test of differential expression hangs on the assuBmption
that these biases are identical between replicates.  Additionally, in many tests
of differential expression---similarly the effects of transcript length---a higher
read count will result in higher statistical confidence. It follows that the
sensitivity of such a test will also be biased by sequence, affecting downstream
systems biology analysis such as gene ontology enrichment tests.

This bias, though observed primarily in the 5' end of a read, is not resolved by
trimming the reads prior to mapping (see, supplementary Figure 1), suggesting
it is not a result of erroneous base calling, and that a more sophisticated
means of correction is needed.

@@

Some work has been done to model and correct for these biases.  Li, et. al.,
\cite{Li2010} propose two models. The firsts is a Poisson linear model, in which
read counts across a transcript follow inhomogeneous Poisson process. The read
count at position $i$ within the transcript is Poisson distributed with
parameter $\lambda_i$, where, $\log(\lambda_i)$ is the sum of weights determined by the
nucleotide at each position surrounding the read start, in addition to a term
capturing the abundance of the transcript.

The second model is based on multiple additive regression trees, or MART
\cite{Friedman2003}.  In their tests, MART shows a moderate improvement over the
Poisson linear model. Both models are fit to a number of abundant test genes,
requiring existing gene annotations for the reference genome.

Another model, proposed by Hansen, et. al., \cite{Hansen2010} directly estimates
the distribution of initial heptamers within reads, then estimates a presumed
background heptamer distribution by sampling from the end of reads. Read counts
are then adjusted by the ratio of the two. Specifying two distributions over
heptamers (i.e. foreground and background distributions) requires $2(4^7-1) =
32,766$ parameters, so while no gene annotations are needed to train such a
model, a significant number of reads are required.

Version 0.9.0 of Cufflinks \cite{Trapnell2010} also introduces a model to
correct for sequence bias. No description of their method is yet available, nor
is there any implementation of the model that can be run independently of
Cufflinks to facilitate testing, so we have withheld any comparison in our
tests.

Besides the direct influence of nucleotide composition on sequencing efficiency,
the ``uniqueness'' of a sequence within the genome affects how likely it is to
be mapped correctly, thus influencing quantification in a non-trivial way. Some
methods exist attempting to quantify uniqueness \cite{Koehler2010} and to take
it into account in the measure of transcript abundance \cite{Li2010b}
\cite{Lee2010}. Though we make no attempt to address this problem, any
probabilistic characterization of uniqueness or ``mapability'' could easily be
combined with the probabilistic characterization of sequence bias that we
described here.

\begin{figure}
\centerline{\includegraphics[width=0.35\textwidth]{fig/overview.eps}}
\caption{An overview of the approach taken: (a) foreground sequences are sampled
from the regions surrounding the starts of mapped reads, (b) background
sequences are sampled by randomly offsetting foreground positions, (c) a
Bayesian network is classifier is trained to descriminate between the set of
sampled foreground and background sequences, (d) and the classifier is evaluated
at each position within a locus, predicting bias.}
\label{fig:overview}
\end{figure}

@@

We assume a simple model of an RNA-Seq experiment: the number of reads $x_i$
aligned to genomic position $i$ is an unbiased estimate of RNA abundance.
Furthermore, we assume reads may be treated as i.i.d. samples. That
is, if $N$ reads are generated, and $m_i$ is the event that a generated read
maps to position $i$,
$$ E[ x_i ] = N \Pr[ m_i ] $$

The experiment may be considered unbiased  with regards to sequence if having
observed the nucleotide sequence $s_i$ surrounding position $i$,
$$ E[ x_i | s_i ] = N \Pr[ m_i | s_i ] = N \Pr[ m_i ] = E[ x_i ] $$

@@

This suggests a natural reweighing scheme. First, define the \emph{sequence
bias} $b_i$ at position $i$ as,
$$ b_i = \frac{ \Pr[ s_i ] } { \Pr[ s_i | m_i ] } $$

Now, if we reweight the read count at position $i$ by $b_i$, we
have,
\begin{align*}
E[ b_i x_i  | s_i ] &= b_i E[ x_i | s_i ] \\
&= N b_i \Pr[ m_i | s_i ] \\
&= N \frac{ \Pr[ m_i | s_i ] \Pr[ s_i ] }{ \Pr[ s_i | m_i ] } \\
&= N \Pr[ m_i ] \\
&= E[ x_i ]
\end{align*}
Thus the reweighted read counts are made to be unbiased.

To estimate the bias $b_i$, we must make estimates of the background sequence
probability $\Pr[s_i]$ and the foreground sequence probability $\Pr[ s_i | m_i
]$, the later being the probability of the sequence given a read being sampled
from its position. Any estimation of sequence probabilities presents a challenge
of finding a model that sufficiently complex to capture the common features of
the training data but avoids overfitting.

Towards that end, we propose training a Bayesian network classifier trained on
examples of foreground and background sequences. Though our goal is not
classification, by using the machinery of classification we can avoid a model
that is over-parametrized. Parameters that are not informative in
discriminating between foreground and background are not included in the model.
The Bayesian network can then used to evaluate sequence probability, and thus
bias at any genomic position.

This discussion so far ignores one complication. The RNA abundance that we wish to
estimate is not itself independent of of the nucleotide sequence. Notably,
exonic DNA is tends to be more GC-rich than intergenic DNA. If background
sequences are sampled uniformly from the genome we run the risk of incorrectly
adjusting for biological sequence bias, rather than technical sequence bias. To
avoid this, we propose using paired training data. Each foreground training
sequence is paired with a background sequence taken from nearby position that is
likely to have similar abundance and general nucleotide composition.
Alternatively, we could pair foreground samples with background samples from
within the same transcript, but we prefer to avoid dependence on existing gene
annotations as such a model would be less suitable for de-novo discovery or
annotation applications.

Hansen, et. al., \cite{Hansen2010} implicitly characterizes and corrects for
bias in the same manner proposed here. Their approach differs primarily in how
sequence probabilities are estimated. Li, et. al., \cite{Li2010} offers a
different perspective, taking advantage of gene annotations to train a model
that optimally explains variance in base level counts across transcripts.

@@

To estimate sequencing bias, we train a Bayesian network classifier in which
each node represents a position in the sequence, relative to the read start, and
edges encode dependency between positions. Bayesian networks have been
applied to recognize motifs in nucleotide sequences in the past, in particular
in modeling splice cites \cite{Cai2000} \cite{Chen2005} and transcription factor
binding sites \cite{Ben-Gal2005} \cite{Grau2006} \cite{Pudimat2005}.

@@

@@

\begin{comment}
\begin{figure*}
\begin{center}
\includegraphics[width=0.75\textwidth]{fig/structure2.pdf}
\end{center}
\caption{The network structure learned on the Trapnell data set, where 0 is the
read start. Dotted boxes represent positions not included in the model (i.e.
positions deemed uninformative).}
\label{fig:structure}
\end{figure*}
\end{comment}

@@

@@

%% TODO
\begin{comment}
\begin{figurehere}
\begin{center}
\includegraphics[width=0.5\textwidth]{fig/kl.pdf}
\end{center}
\caption{Plots of the KL divergence over overlapping $k$-mers in the Trapnell
experiment.}
\label{fig:kl}
\end{figurehere}
\end{comment}

@@

%% TODO
\begin{comment}
\begin{figurehere}
\begin{center}
\includegraphics[width=0.4\textwidth]{fig/pois.pdf}
\end{center}
\caption{Plotted is the change in fit of each of the test exons. A
positive value indicates a better fit in the bias corrected model.}
\label{fig:pois}
\end{figurehere}
\end{comment}

>>>>

\section{Motivation:}
Quantification of sequence abundance in RNA-Seq and ChiP-Seq experiments is
often conflated with protocal-specific bias, influenced by PCR amplification,
and differing primer affinities and mixtures, for example. The result is
potentially inaccurate quantification, affecting de novo gene annotation and
isoform quantification in RNA-Seq, and peak calling in ChiP-Seq.

\section{Results:}
We present a method to measure and correct for these influences using a simple
graphical model. Our model does not rely on existing gene annotations, making it
applicable to any high throughput sequencing data. We evaluate our method on
several datasets, and by multiple criteria, demonstrating that it effectively
decreases bias and increases uniformity.

@@

A significant example is the effect of transcript length on quantification.  As
noted by \citet{Oshlack2009} transcript length biases discovery of
differentially expression, since more reads are aligned to longer transcripts,
granting higher statistical significance in tests differential expression.
Recently, some methods have been developed attempting to account for this effect
in gene set enrichment tests \citep{Young2010,Gao2011}.

%% TODO: do something with this sentance.
Additionally,
differing transcript lengths introduce ``edge effects'', as very short fragments
from the end of a transcript will not be sequenced. NEUMA \cite{Lee2010} and
FPKM \cite{Trapnell2010} are two methods that attempt to take this into account.

The underlying nucleotide sequence also influences quantification in a manner
that is complex and protocol-specific. In an ideal experiment, the number of
reads mapping to a particular position in the genome is a function of RNA
abundance and should not be additionally dependent on the sequence at that
position. As noted first by \citet{Dohm2008}, this is not the case.  As
illustration, Figure \ref{fig:freqs} plots this nonuniformity in nucleotide
frequencies on five data sets (see, Table \ref{tab:datasets}), each using a
different protocol.

These biases may adversely effect transcript discovery, as low level noise may
be over reported in some regions, and in others, active transcription may be
under reported. They render untrustworthy comparisons of relative abundance
between genes or isoforms, and any test of differential expression hangs on the
assumption that these biases are identical between replicates.  Additionally, in
many tests of differential expression---similar to the effects of transcript
length---a higher read count will result in higher statistical confidence. It
follows that the sensitivity of such a test will also be biased by sequence,
affecting downstream analysis such as gene ontology enrichment tests.

This bias, though observed primarily in the 5' end of a read, is not resolved by
trimming the reads prior to mapping \citep{Hansen2010} (see also, Supplementary
Section 1), suggesting it is not a result of erroneous base calling, and that a
more sophisticated means of correction is needed.

@@

Li, et. al., \cite{Li2010} propose two models. The first is a Poisson linear
model, in which read counts across a transcript follow inhomogeneous Poisson
process. The read count at position $i$ within the transcript is Poisson
distributed with parameter $\lambda_i$, where, $\log(\lambda_i)$ is the sum of
weights determined by the nucleotide at each position surrounding the read
start, in addition to a term capturing the abundance of the transcript.

The second model is based on multiple additive regression trees, or MART
\citep{Friedman2003}.  In their tests, the MART model shows a moderate
improvement over the Poisson linear model. Both models are fit to a number of
abundant test genes, requiring existing gene annotations for the reference
genome.

Another model, proposed by \citet{Hansen2010}, directly estimates the
distribution of initial heptamers within reads, then estimates a presumed
background heptamer distribution, sampled from the end of reads. The read count
at a given position is then adjusted by the ratio of the foreground and
background heptamer probabilities. Specifying two distributions over heptamers
(i.e. foreground and background distributions) requires $2(4^7-1) = 32,766$
parameters, so while no gene annotations are needed to train such a model, a
significant number of reads are required.

Lastly, \citet{Roberts2011} have very recently published a description of
another approach, in which sequence probabilities are modeled by variable-order
Markov chains. The structure of these Markov chains are hard-coded, chosen in
advance using a hill-climbing algorithm on a representative dataset. This method
is implemented in the latest version of Cufflinks \citep{Trapnell2010}, and
tightly incorporated into its estimation of transcript abundance.

Here we propose a new approach, using Bayesian networks to model sequence
probabilities. Unlike the methods of Roberts or Li, our model requires no gene
annotations, nor even the assumption that the short reads are derived from RNA.
In this sense, we build on the work done by \citet{Hansen2010}, generalizing
their approach in a way we find to be more robust and effective at correcting
for bias in a variety of protocols. Due to the weak assumptions required my our
model, it is applicable and potentially useful in any setting in which short
reads are aligned to a reference sequence.

%% TODO: where should this figure be cited?
\begin{figure}
\centerline{\includegraphics[width=0.35\textwidth]{fig/overview.eps}}
\caption{An overview of the approach taken: (a) foreground sequences are sampled
from the regions surrounding the starts of mapped reads, (b) background
sequences are sampled by randomly offsetting foreground positions, (c) a
Bayesian network is classifier is trained to descriminate between the set of
sampled foreground and background sequences, (d) and the classifier is evaluated
at each position within a locus, predicting bias.}
\label{fig:overview}
\end{figure}

@@

We begin with a natural model of an RNA-Seq experiment (and one that is often
assumed, whether implicitly or otherwise). The number of reads $x_i$ aligned to
genomic position $i$ is an unbiased estimate of RNA abundance. Furthermore, we
assume reads may be treated as independent and identically distributed samples.
That is, if $N$ reads are generated, and $m_i$ is the event that a generated
read maps to position $i$.
$$ E[x_i] = N \Pr[m_i] $$

The experiment may be considered unbiased  with regards to sequence if, having
observed the nucleotide sequence $s_i$ surrounding position $i$,
$$ E[ x_i | s_i ] = N \Pr[ m_i | s_i ] = N \Pr[ m_i ] = E[ x_i ] $$

@@

This suggests a natural scheme in which observations may be reweighted to
correct for bias.  First, define the \emph{sequence bias} $b_i$ at position $i$
as,
$$ b_i = \frac{ \Pr[ s_i ] } { \Pr[ s_i | m_i ] } $$

Now, if we reweight the read count $x_i$ at position $i$ by $b_i$, we
have,
\begin{align*}
E[ b_i x_i | s_i ] &= b_i E[ x_i | s_i ] \\
&= N b_i \Pr[ m_i | s_i ] \\
&= N \frac{ \Pr[ m_i | s_i ] \Pr[ s_i ] }{ \Pr[ s_i | m_i ] } \\
&= N \Pr[ m_i ] \\
&= E[ x_i ]
\end{align*}
Thus the reweighted read counts are made unbiased.

To estimate the bias $b_i$, we must make estimates of the background sequence
probability $\Pr[s_i]$ and the foreground sequence probability $\Pr[ s_i | m_i
]$, the latter being the probability of the sequence given a read being sampled
from its position. Estimating bias is therefore a problem of finding a model of
sequence probability that sufficiently complex to capture the common features of
the training data yet avoids overfitting.

Towards that end, we propose training a Bayesian network classifier trained on
examples of foreground and background sequences. Though our goal is not
classification, by using the machinery of classification we can avoid a model
that is over-parametrized. Parameters that are not informative in
discriminating between foreground and background are not included in the model.
The Bayesian network can then used to evaluate sequence probability, and thus
bias, at any genomic position.

However, we have so for ignored one complication: the RNA abundance that we wish
to estimate is not itself independent of the nucleotide sequence. Notably,
exonic DNA tends to be more GC-rich than intergenic DNA. If background sequences
are sampled uniformly from the genome we run the risk of incorrectly adjusting
for biological sequence bias, rather than techncical sequence bias.  To avoid
this, we propose using paired training data. Each foreground training
sequence is paired with a background sequence taken from nearby position
that is likely to have similar abundance and general nucleotide composition.
Alternatively, we could pair foreground samples with background samples from
within the same transcript, but we prefer to avoid dependence on existing
gene annotations as such a model would be limited to RNA-Seq analysis in which
trustworthy and reasonably complete gene annotations are available.

The methods proposed \citet{Hansen2010} and \cite{Roberts2011} also treat bias
correction as a problem of estimating foreground and background sequence
probabilities. They differ primarily in how these sequence probabilities are
estimated. \citet{Li2010} estimate reweighting coefficients ($b_i$, in our
notation) directly, given training data consistig of long, highly-expressed
transcripts.

@@

To estimate sequencing bias, we train a Bayesian network classifier in which
each node represents a position in the sequence, relative to the read start, and
edges encode dependency between positions.  Bayesian networks have been applied
to recognize motifs in nucleotide sequences in the past, in particular in
modeling splice cites \citep{Cai2000, Chen2005} and transcription factor binding
sites \citep{Ben-Gal2005, Grau2006, Pudimat2005}.

@@

%% TODO: leaving it here for now
%% -------------------------------------------------------

@@

@@

\begin{figure*}
\centerline{\includegraphics[width=\textwidth]{fig/models.eps}}
\caption{Blah blah blah.}
\label{fig:models}
\end{figure*}

@@

\begin{figure*}
\centerline{\includegraphics[width=0.9\textwidth]{fig/kl.eps}}
\caption{Plots of the KL divergence over overlapping $k$-mers in the Trapnell
experiment.}
\label{fig:kl}
\end{figure*}

@@

%% TODO
\begin{figure}
\centerline{\includegraphics[width=0.3\textwidth]{fig/pois-boxplot.eps}}
\caption{Plotted is the change in fit of each of the test exons. A
positive value indicates a better fit in the bias corrected model.}
\label{fig:pois}
\end{figure}
